Statistical dynamics of SARS-CoV-2 as a differential game

### 2020-07-23 paper/paper.tex
23JUL proof version.

\begin{abstract}
    Abstract
\end{abstract}

\section{Introduction} % (fold)
\label{sec:introduction}
Where an infectious disease is not amenable to population-level prevention through vaccination and risks are non-trivial, non-pharmaceutical interventions (NPIs) remain the principal tool of public health to respond to an outbreak. This is the case with novel infectious diseases that have no specific treatment and no prophylactic (vaccine) available. In the absence of pharmaceutical interventions of proven effectiveness, in particular prophylactically, the main public health response to the emerging pandemic of COVID-19, a viral syndrome caused by the (+)ssRNA virus SARS-CoV-2 (order \emph{Nidovirales}, family \emph{Coronaviridae}, genus \emph{Betacoronavirus}, subgenus \emph{Sarbecovirus}), has rested principally on NPIs, first and foremost social distancing and ancillary steps intended to facilitate that.

In general, any course of conduct that reduces the encounter rate between an individual and other individuals can be considered a form of social distancing. This may be brought about through limiting public facilities for such encounters ('lockdowns'), through limiting individual gatherings by size ('large-gathering bans') and through encouraging individual social distancing. From the perspective of game theory, social distancing can be viewed as a non-cooperative game of a population $P_{1 \ldot n}$ of size $n$, where at any given time $t \in [t_0, t_{e}]$, the strategy adopted by $p_i$ is denoted as $\sigma (p_i, t)$. For the sake of simplicity, we assume that distancing is either not exercised at all or perfectly exercised, i.e. $\sigma (p_i, t) \in \{ 0, 1 \}$. Then, for the entire population, the overall strategy can be described as

\begin{equation}
	\bar{\sigma}(P, t) = \frac{\displaystyle \sum_{i = 0}^n \sigma(p_i, t)}{n}
	\label{eq:overall_strategy}
\end{equation}

\noindent and for the entire time period, from $t_0$ to the endpoint $t_e$ (which may be eradication, elimination, natural extinction of the pathogen, the availability of a vaccine or a combination thereof), for discrete time $t$,

\begin{equation}
	\bar{\bar{\sigma}}(P) = \frac{\displaystyle \sum_{i = 0}^n \displaystyle \sum_{j = 0}^{t_e - t_0} \sigma(p_i, j)}{n (t_e - t_0)}
	\label{eq:overall_strategy_over_time}
\end{equation}

However, with each course of conduct, there is associated a cost $J(\sigma)$. For simplicity's sake, let us consider these costs to be governed by the following three precepts given a pathogen with fixed characteristics ($R_0$, transmission potential, risks \&c.). Let $\bar{\delta}(P)$ equal the proportion of persons $p \in P$ adopting strategy $\sigma_{\delta}$. It then holds that:

\begin{enumerate}
	\item A person $p_i \in P$ opting for strategy $\delta$ (social distancing) will incur $c_d$, the immediate costs of distancing. These may be social (lessened social interaction), psychological (lessened access to support systems), economic (lower access to facilities to earn) or simple matters of convenience (access to amenities). While $c_d$ is somewhat dependent on $\bar{\bar{(P)}}$ (thus not distancing does not yield a benefit to a lone social distancer in terms of access to amenities if all of the latter are closed), it can be assumed to be largely constant.
	\item Compared to a person opting for strategy $\delta$, a person $p_i \in P$ opting for strategy $\lnot \delta$ will incur a relative additional risk $f_r(\bar{\delta}(P), t)$, i.e. contingent on the population's behaviour at time $t$.
	\item Finally, a person $p_i \in P$ will regardless of their individual choice (at least at non-trivial population sizes) gain a protective benefit, which is a function of the population's adherence to social distancing. We will denote this term as $f_c(\bar{\delta}(P), t)$.
\end{enumerate}

Then, assigning the variable $c_s$ to represent the cost of illness (economic loss, medical costs, long-term health risks), we can express the cost functions of each strategy, $\sigma_{\delta}$ (social distancing) and $\sigma_{\lnot \delta}$ (no social distancing), as

\begin{equation}
	J_{\sigma_{\delta}}(p_i, t) = c_d - f_c(\bar{\delta}(P), t)
\end{equation}

@@

\begin{equation}
	J_{\sigma_{\lnot \delta}}(p_i, t) = f_r(\bar{\delta}(P), t) c_s - f_c(\bar{\delta}(P), t)
	\label{eq:risk}
\end{equation}

\noindent And consequently, it holds for the population level cost $\bar{J}$ at time $t$ that

\begin{equation}
	\bar{J}(P, t) = \frac{\displaystyle \sum_{i=0}^{\delta(P) n} J_{\sigma_{\delta}}(p_i, t) + \displaystyle \sum_{j=0}^{1-\delta(P) n} J_{\sigma_{\lnot \delta}}(p_j, t)}{n}
	\label{eq:cost_eqn}
\end{equation}

\noindent and for a population over time $t_0$ to $t_e$,

\begin{equation}
	\bar{\bar{J}}(P) = \int_{t_0}^{t_e} \bar{J}(P, t) dt
\end{equation}

As Reluga (2010) notes, the effect of social distancing diminishes with the increase in participants, the phenomenon of diminishing returns.\cite{reluga2010game} Thus, as the number of individuals in $P$ opting for social distancing increases, the marginal increase by another social distancer diminishes. This can be represented by a discount factor $h$, to yield

\begin{equation}
	\bar{J}(P) = \int_{t_0}^{t_e} e^{-ht} \frac{\delta(P) J_{\sigma_{\delta}}(P, t) + (1 - \delta(P)) J_{\sigma_{\lnot \delta}}(P, t)}{n}
\end{equation}

However, because individual decisions affect the overall gain (due to the component dependent on $\delta(P, t)$), our principal concern is not with individual action but with analysis of such strategies on a population level. This paper will in the following conceptualise infectious disease in a population as a differential game over a differential equation form of the compartmental model first described by Kermack and McKendrick\cite{kermack1927contribution} and, since its publication in 1927, widely adapted and adopted.\cite{vstvepan2007kermack,roberts1999kermack,capasso1978generalization}

@@

\begin{equation}
	\begin{aligned}
		\frac{dS}{dt} = - \frac{\beta S I}{n} 								\\
		\frac{dI}{dt} = \frac{\beta S I}{n} - \gamma I 						\\
		\frac{dR}{dt} = \gamma I
	\end{aligned}
	\label{eq:sir_equation}
\end{equation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/fig1-odes}
	\caption{Some quantitative solutions for the SIR model's population dynamics over values of $R_0$ between 1.5 and 6.5, and values of $\gamma$ between $1/4.6$ and $1/12.4$ over a base population of 10,000 and a seed population of 0.1\% infected initially. For each plot, $\beta$ is inferred from an $R_0$ value of 2.67, based on Liu et al. (2020),\cite{liu2020reproductive} and the $\gamma$ parameter. The susceptible population is displayed in blue, while infected/infectious cases are marked in red and recovered cases in green.}
	\label{fig:ode_solutions}
\end{figure}

\noindent under the assumption of $S + I + R = 0$, where $S$ represents susceptible individuals, $I$ represents infected/infectious individuals and $R$ accounts for removed individuals (mortality and recovery to immunity). The fraction $\frac{\beta}{\gamma}$ equals the basic reproduction number, $R_0$. For SARS-CoV-2, estimates of $R_0$ range from 1.4 to 6.49, with studies that relied on statistical estimation of $R_0$ ranging from 2.20 to 3.58, with an average of 2.67\cite{liu2020reproductive} $\gamma$, on the other hand, can be estimated as the invers of the number of days of illness, which can be approximated as $8.5 \pm 3.9$ days.\cite{pan2020clinical,liu2020risk} Figure~\ref{fig:ode_solutions} describes some analytical solutions for the differential equations of Equation~\eqref{eq:sir_equation} over a range of plausible values of $R_0$ and $\gamma$ under the assumption of an $R_0$ of 2.67. Even in the absence of firm evidence as to whether SARS-CoV-2 infection followed by recovery would engender lifelong immunity or not,\cite{roy2020covid,ota2020will,lin2020duration} it can be assumed in the short term -- based on evidence from MERS-CoV and SARS-CoV -- that in the short term, survivors remain immune,\cite{prompetchara2020immune} and consequently the number of individuals in $R$ does not decrease. Numerical solutions to this system of differential equations have been calculated using \texttt{odepack} via \texttt{SciPy 1.5.1}\cite{virtanen2020scipy} on Python 3.6, and are presented as Figure~\ref{fig:ode_solutions} for a range of values of $R_0$ and $\gamma$.

@@

\begin{equation}
	\begin{aligned}
		\frac{dS}{dt} = - \frac{\beta S I - \delta \beta S I}{n} 			\\
		\frac{dI}{dt} = \frac{\beta S I - \delta \beta S I}{n} - \gamma I	\\
		\frac{dR}{dt} = \gamma I
	\end{aligned}
	\label{eq:sir_with_social_distancing}
\end{equation}

@@

While quantification of costs of illness is difficult, quantification of the economic, social and emotional costs of social distancing is even harder to grasp. However, we may, for a range of cost fractions $\frac{c_d}{c_i}$, estimate values of $\delta(t)$ that denote optimal strategic spaces. Thus,

\begin{equation}
	\begin{aligned}
		\frac{c_d}{c_i}(P, t) = \frac{(1 - \delta(t)) R_0 \gamma S(t) I(t)}{\delta(t) n^2}
	\end{aligned}
	\label{eq:cost_fraction}
\end{equation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/cost_fraction}
	\caption{Cost fraction $\frac{c_d}{c_i}$ of social distancing in a population $P$ of $\delta(P)$ adherence over time, based on a population of 10,000 individuals with a seed population of 0.1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$. The contour lines indicate the cost fraction, i.e. what fraction of the cost of social distancing $c_d$ the cost of illness $c_i$ must be in order to make not distancing a preferred strategy.}
	\label{fig:cost_fraction}
\end{figure}

As numerical estimation of this cost fraction (Figure~\ref{fig:cost_fraction}) shows that for most cases, not distancing can only be a feasible strategy if the cost of distancing very significantly outweighs the cost of illness, often by as much as four orders of magnitude. Since this is notwithstanding risk (i.e. the comparison needs to be made between the absolute cost of infection and the absolute cost of distancing, rather than the probability/risk-adjusted figures).

@@

@@

>>>>

\begin{abstract}
    The novel coronavirus SARS-CoV-2 has rapidly emerged as a significant worldwide threat to public health, in particular because -- as is not uncommon with novel pathogens -- there is no effective pharmaceutical treatment or prophylaxis to the viral syndrome it causes. In the absence of such specific treatment modalities, the mainstay of public health response rests on non-pharmaceutical interventions (NPIs), such as social distancing. This paper contributes to the understanding of social distancing against SARS-CoV-2 by quantitatively analysing the statistical dynamics of disease propagation as a differential game, and estimating the relative costs of distancing versus not distancing, identifying marginal utility of distancing based on known population epidemiological data about SARS-CoV-2 and concluding that unless the costs of distancing vastly exceed the cost of illness per unit time, social distancing remains a dominant strategy. These findings can assist in solidly anchoring public health responses based on social distancing within a quantitative framework attesting to their effectiveness.
\end{abstract}

\section{Introduction} % (fold)
\label{sec:introduction}
Where an infectious disease is not amenable to population-level prevention through vaccination and risks are non-trivial, non-pharmaceutical interventions (NPIs) remain the principal tool of public health to respond to an outbreak. This is the case with novel infectious diseases that have no specific treatment and no prophylactic (vaccine) available. In the absence of pharmaceutical interventions of proven effectiveness, in particular prophylactically, the main public health response to the emerging pandemic of COVID-19, a viral syndrome caused by the (+)ssRNA virus SARS-CoV-2 (order \emph{Nidovirales}, family \emph{Coronaviridae}, genus \emph{Betacoronavirus}, subgenus \emph{Sarbecovirus}), has rested principally on NPIs.\cite{mccoy2020can,lai2020effect,flaxman2020report,ferguson2020report} At their core, all NPIs share a quintessential relationship to social distancing, either through directly encouraging social distancing, limiting encounters by reducing public facilities for such encounters that may transmit the pathogen ('lockdowns'), reducing gatherings and social interactions that carry such risk ('large-gathering bans'), suspending economic activities that inherently carry the risk of social interactions or modifying the framework of such activities to reduce such interactions (e.g. transitioning to remote work).

From the perspective of game theory, social distancing can be viewed as a non-cooperative game of a population $P_{1 \ldots n}$ of size $n$, where at any given time $t \in [t_0, t_f]$, each player $p_i$ adopts the strategy $\sigma(p_i, t)$. For the sake of simplicity, we will assume two fundamental properties of social distancing conduct:

\begin{enumerate}
	\item There exist two strategy choices, $\delta$ and $\lnot \delta$ (distancing and not distancing, respectively). At any given time, any agent $p_i \in P$ can opt exclusively for one of these two options, i.e. the strategy set $\mathcal{S}(p_i)$ is the set $\{\delta, \lnot \delta\}$.
	\item Where an agent $p_i$ makes a policy choice at time $t$, $\sigma(p_i, t) \in [\delta, \lnot \delta]$, they implement it perfectly, i.e. the efficiency of every actor $p_i \in P$ in implementing their strategy $\sigma(p_i, t)$ is equal.
\end{enumerate}

Then, for the entire population $P$, the overall strategy $\bar{\sigma}(P, t)$ can be described as the sum of all strategies $\sigma(p_i, t) | p_i \in P$. We can then assign a value $\hat{\sigma}$ to each strategy, whereby the strategy of distancing, $\delta$, is assigned the value of 1 and the strategy of not distancing, $\lnot \delta$, the value of 0.

For any given discrete time $t \in [t_0, t_f]$, where $t_f$ would be an endpoint (such as eradication, elimination, natural extinction of the pathogen, the availability of a vaccine or a combination thereof), we may then define two disjoint sets, $P_{\delta}(t)$ and $P_{\lnot \delta} (t)$, where

\begin{equation}
	\forall p_j \in P_{\delta}(t), \ \ \sigma(p_j, t) = \delta
\end{equation}

@@

\begin{equation}
	\forall p_k \in P_{\lnot \delta}(t), \ \ \sigma(p_k, t) = \lnot \delta
\end{equation}

\noindent so that $|P_{\delta}(t)| + |P_{\lnot \delta}(t)| = |P| = n$. Then, for any $t$ in discrete time $t \in [t_0, t_f]$, we define the overall factor of social distancing for a population $P$ at time $t$, $\delta(P, t)$

\begin{equation}
	\begin{aligned}
		\delta(P, t) = \frac{P_{\delta}(t)}{n} = 1 - \frac{P_{\lnot \delta}(t)}{n}
	\end{aligned}
\end{equation}

\noindent which can also be expressed as a function of strategies $\sigma(p_{1 \ldots n})$ as

\begin{equation}
	\begin{aligned}
		\delta(P, t) = \frac{\sum_{i=1}^n \hat{\sigma}(p_i, t)}{n}
	\end{aligned}
	\label{eq:delta_value}
\end{equation}

In other words, the overall factor of social distancing at time $t$ is the average of the value assigned to each agent's strategy over the entire population $P$. For simplicity's sake, we will assume that the decision process takes the shape of a continuous and simultaneous game in discrete time, and agents can instantly switch strategies with no cost (other than the cost of the strategy itself). It then holds that

\begin{enumerate}
	\item A person $p_i \in P$ opting for strategy $\sigma_{\delta}$ (social distancing) will incur $c_d$, the immediate costs of distancing. These may be social (lessened social interaction), psychological (lessened access to support systems), economic (lower access to facilities to earn) or simple matters of convenience (access to amenities). While $c_d$ is somewhat dependent on $\delta(P, t)$ (thus not distancing does not yield a benefit to a lone social distancer in terms of access to amenities if all of the latter are closed), it can be assumed to be largely constant.
	\item Compared to a person opting for strategy $\sigma_{\delta}$, a person $p_i \in P$ opting for strategy $\sigma_{\lnot \delta}$ will incur a relative additional risk $r_i(P, t) c_i$, where $c_i$ denotes the constant cost of illness and $r_i(P, t)$ is the risk of contracting illness when not distancing, a function of $\delta(P, t)$.
\end{enumerate}

This allows us to identify the cost function $J(\sigma, t)$ for any $\sigma \in \{\delta, \lnot \delta\}$ at time $t$ for each individual $p_i \in P$ as

\begin{equation}
	\begin{aligned}
		J(\sigma_{\delta}, t) = c_d 									\\
		J(\sigma_{\lnot \delta}, t) = r_i(P, t) c_i
	\end{aligned}
	\label{eq:pure_time_dependent_costs}
\end{equation}

Then, for every social strategy $\bar{\sigma}(P, t)$ associated with the social distancing value $\delta(P, t)$ in the way described in Equation~\eqref{eq:delta_value}, the overall social cost of $\bar{\sigma}(P, t)$ can be conceptualised as

\begin{equation}
	\begin{aligned}
		\bar{J}(P, t) = n \Big( \delta(P, t) c_d + (1-\delta(P, t)) r_i(P, t) c_i \Big)
	\end{aligned}
	\label{eq:social_time_dependent_costs}
\end{equation}

For the entire time space $[t_0, t_f]$, the total social cost is then

\begin{equation}
	\bar{\bar{J}}(P) = \sum_{t_i = 0}^{t_f - t_0} J(P, t_i) = n \sum_{t_i = 0}^{t_f - t_0} \Big( \delta(P, t_i) c_d + (1 - \delta(P, t_i)) r_i(P, t_i) c_i \Big)
\end{equation}

Our principal concern in this paper is not with individual action but with analysis of decision strategies on a population level. This paper will in the following conceptualise infectious disease in a population as a differential game over a differential equation form of the compartmental model first described by Kermack and McKendrick\cite{kermack1927contribution} and, since its publication in 1927, widely adapted and adopted,\cite{vstvepan2007kermack,roberts1999kermack,capasso1978generalization} and identify within the statistical dynamics of that differential game the equilibria that govern ideal societal decision-making.

@@

\begin{equation}
	\begin{aligned}
		\frac{dS}{dt} &= - \frac{\beta S I}{n} 								\\
		\frac{dI}{dt} &= \frac{\beta S I}{n} - \gamma I 					\\
		\frac{dR}{dt} &= \gamma I
	\end{aligned}
	\label{eq:sir_equation}
\end{equation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/fig1-odes}
	\caption{Some quantitative solutions for the SIR model's population dynamics over values of $R_0$ between 1.5 and 6.5, and values of $\gamma^{-1}$ between $4.6$ and $12.4$ over a base population of 10,000 and a seed population of 0.1\% infected initially. For each plot, $\beta$ is inferred from an $R_0$ value of 2.67, based on Liu et al. (2020),\cite{liu2020reproductive} and the $\gamma$ parameter. The susceptible population is displayed in blue, while infected/infectious cases are marked in red and recovered cases in green.}
	\label{fig:ode_solutions}
\end{figure}

\noindent under the assumption of $S + I + R = 0$, where $S$ represents susceptible individuals, $I$ represents infected/infectious individuals and $R$ accounts for removed individuals (mortality and recovery to immunity). The fraction $\frac{\beta}{\gamma}$ equals the basic reproduction number, $R_0$. For SARS-CoV-2, estimates of $R_0$ range from 1.4 to 6.49, with studies that relied on statistical estimation of $R_0$ ranging from 2.20 to 3.58, with an average of 2.67\cite{liu2020reproductive} $\gamma$, on the other hand, can be estimated as the inverse of the average number of days of illness ($\gamma^{-1}$, sometimes also described as $\tau$), which can be approximated as $8.5 \pm 3.9$ days.\cite{pan2020clinical,liu2020risk} Figure~\ref{fig:ode_solutions} describes some analytical solutions for the differential equations of Equation~\eqref{eq:sir_equation} over a range of plausible values of $R_0$ and $\gamma$ under the assumption of an $R_0$ of 2.67. Even in the absence of firm evidence as to whether SARS-CoV-2 infection followed by recovery would engender lifelong immunity or not,\cite{roy2020covid,ota2020will,lin2020duration} it can be assumed in the short term -- based on evidence from MERS-CoV and SARS-CoV -- that in the short term, survivors remain immune,\cite{prompetchara2020immune} and consequently the number of individuals in $R$ does not decrease. Numerical solutions to this system of differential equations have been calculated using \texttt{odepack} via \texttt{SciPy 1.5.1}\cite{virtanen2020scipy} on Python 3.6, and are presented as Figure~\ref{fig:ode_solutions} for a range of values of $R_0$ and $\gamma$.

@@

\begin{equation}
	\begin{aligned}
		\frac{dS}{dt} &= - \frac{\beta S I - \delta \beta S I}{n} 			\\
		\frac{dI}{dt} &= \frac{\beta S I - \delta \beta S I}{n} - \gamma I	\\
		\frac{dR}{dt} &= \gamma I
	\end{aligned}
	\label{eq:sir_with_social_distancing}
\end{equation}

@@

While quantification of costs of illness is difficult, quantification of the economic, social and emotional costs of social distancing is even harder to grasp. However, we may, values of $\delta(P)$, calculate cost fractions $\phi(P, t) = \frac{c_d}{c_i}(P, t)$, where

\begin{equation}
	\begin{aligned}
		\phi(P, t) = \frac{c_d}{c_i}(P, t) = \frac{(1 - \delta(t)) R_0 \gamma S(t) I(t)}{\delta(t) n^2}
	\end{aligned}
	\label{eq:cost_fraction}
\end{equation}

This cost fraction indicates the relative disequilibrium between the cost of distancing and the cost of illness when adjusted for risk -- in other words, for any given value of $\delta(P, t)$ at $t$, the cost fraction indicates an inflection point. As long as the cost of distancing is less than $\phi c_i$, social distancing at $\delta(P, t)$ or above is the optimum strategy. As numerical estimation of this cost fraction (Figure~\ref{fig:cost_fraction}) shows, social distancing is almost always the preferred strategy at equal cost (black line). The contour lines in Figure~\ref{fig:cost_fraction} indicate what ratio the cost of distancing has to be to the cost of illness at a given fraction $\delta(P, t)$ of distancing to make distancing no longer an optimal strategy. So, for instance, a $\phi$ of 0.05 denotes the isorisk curve over $t$ where, at given $\delta(P, t)$, distancing is the preferable solution as long as its costs are 0.05 times that of illness or less.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/cost_fraction}
	\caption{Cost fraction $\phi = \frac{c_d}{c_i}$ of social distancing in a population $P$ of $\delta(P, t)$ adherence over time, based on a population of 10,000 individuals with a seed population of 0.1\% infected, under the assumption of an $R_0$ of 2.67 and $\gamma$ of $\frac{1}{8.5}$. The contour lines indicate the cost fraction, i.e. what fraction of the cost of social distancing $c_d$ the cost of illness $c_i$ must be in order to make not distancing a preferred strategy.}
	\label{fig:cost_fraction}
\end{figure}

@@

Estimates of direct medical costs of COVID-19 are difficult, but at least one figure puts the median cost of a symptomatic infection at US\$3,045 in direct costs alone,\cite{bartsch2020potential} typically compounded by loss of earnings, long-term physical harm, reduction in life expectancy and quality of life and, in severe cases, the risk of mortality. The costs of social distancing are much less amenable to quantification, as these costs are primarily governed by indirect factors and intangibles, such as the cost of deferred medical treatment, second-order effects of the psychological burden inherent in decreased social interaction and the cost of lost revenue. While quantification, thus, of both the cost of distancing and cost of illness remains an outstanding subject of research, the cost fraction calculations can assist in reasoning about the best social strategy in view of these factors once ascertained.
